---
title: "Info 411 Project - General Data Processing"
output: pdf_document
---

# General Data Processing

#### Setup & Configuration

Loading the libraries and the data set, setting the seed, and defining thresholds + sampling information.

```{r}
library(readxl)
library(dplyr)
library(lubridate)
library(writexl)

set.seed(123)

#Targeting 1000 unique customers
sample_n_customers <- 1000

#Churn thresholds (provided in project brief)
recency_threshold_days <- 90
low_frequency_threshold <- 3
low_monetrary_threshold <- 100

file_path <- "Online Retail.xlsx"
df_raw <- read_excel(file_path)
head(df_raw)
```

#### Cleaning the Data

Removing invalid rows, creating a TotalPrice and Date column and normalizing the data types.

```{r}

#Normalizing the data and creating TotalPrice + Date.
df_clean <- df_raw %>%
  mutate(
    CustomerID = as.character(CustomerID),
    InvoiceNo = as.character(InvoiceNo),
    InvoiceDate = as.POSIXct(InvoiceDate, tz="UTC"),
    Date = as.Date(InvoiceDate),
    Quantity = as.numeric(Quantity),
    UnitPricce = as.numeric(UnitPrice),
    TotalPrice = Quantity * UnitPrice
  ) %>%
  
  #Remove rows with no customerID
  filter(!is.na(CustomerID), CustomerID != "") %>%
  
  #Remove rows with no price or quantity
  filter(Quantity > 0, UnitPrice > 0) %>%
  
  #Remove rows with no date
  filter(!is.na(Date))
  
cat("Rows before cleaning:", nrow(df_raw), "\n")
cat("Rows after cleaning:", nrow(df_clean), "\n")
cat("Rows removed:", nrow(df_raw) - nrow(df_clean), "\n")
```

#### Customer-Level Sampling

Sample unique customers and keep all their transactions, ensuring customers behavioral history is kept.

```{r}
#Calcualte number of unqiue customers
all_customers <- unique(df_clean$CustomerID)
cat("Number of unique customers:", length(all_customers), "\n")

#Sample 1000 customers from dataset
sampled_customers <- sample(all_customers, size = sample_n_customers, replace = FALSE)

df_txn <- df_clean %>%
  filter(CustomerID %in% sampled_customers)

cat("Number of rwas in sampled data:", nrow(df_txn), "\n")
```

#### Create Customer Behavior Features

Converting transnational data into one row per customer, creating usable features for all models.

```{r}
#Getting the most recent date in the dataset (e.g. dataset end date)
analysis_end_date <- max(df_txn$Date, na.rm = TRUE)

#Creating aggregated dataset with new features
df_master <- df_txn %>%
  group_by(CustomerID) %>%
  summarise(
    LastPurchaseDate = max(Date, na.rm = TRUE),
    RecencyDays = as.integer(analysis_end_date - LastPurchaseDate),
    Frequency = n_distinct(InvoiceNo),
    Monetary = sum(TotalPrice, na.rm = TRUE),
    AvgBasketValue = Monetary / pmax(Frequency, 1),
    NumItems = sum(Quantity, na.rm = TRUE),
    Countries = n_distinct(Country),
    .groups = "drop"
  )

head(df_master)
```

#### Handling Extreme Outliers

Remove very high spenders, to improve cluster quality.

```{r}
#Setting maximum total spend, calcualtes the 99th percentile.
cap <- quantile(df_master$Monetary, 0.99, na.rm = TRUE)

#Removing all rows above this value
df_master <- df_master %>%
  filter(Monetary <= cap)

summary(df_master$Monetary)
```

#### Creating Churn Label

add churn label based on project guidelines

```{r}
#Setting 1 or 0 for IsChurned depending on condition
df_master <- df_master %>%
  mutate(
    IsChurned = if_else(
      RecencyDays > recency_threshold_days | (Frequency <= low_frequency_threshold & Monetary <= low_monetrary_threshold),
      1L, 0L
    )
  )

#How many customers fall into each class
table(df_master$IsChurned)
```

#### Final Check

```{r}
summary(df_master)
head(df_master)
```

#### Exporting Processed Dataset

```{r}
write_xlsx(
  list(MASTER = df_master),
  "INFO411_master_preprocessed.xlsx"
)
```
